---
title: "Popularity of Hearthstone Cards"
output: 
    html_document:
      code_folding: show
      df_print: kable
      toc: yes
      toc_float:
        collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following packages are used in this analysis:

```{r libraries, message=FALSE}
library(dplyr)
library(ggplot2)
library(jsonlite)
library(knitr)
library(readr)
library(stringr)
library(tidyr)
library(zoo)
```


## Introduction

Hearthstone is a popular collectible card game published by Blizzard Entertainment in 2014, which is based on the Warcraft series by the same company. The goal of the game is to build a deck of 30 cards and defeat the opponent who also has a deck of 30 cards.

Cards can be classified according to the following categories:

* **Class**: Neutral cards can be used by all nine classes, while Class cards can only be used by the indicated class.
* **Rarity**: How often cards can be found when opening card packs.
    * Cards are indicated as Free, Common, Rare, Epic and Legendary, ordered by increasing rarity.
* **Type**: Different types of cards have different effects in the game:
    * Minions are played on the game board and can attack Heroes or other minions.
    * Spells are Class abilities that generate a variety of effects on the board.
    * Weapons are items that Heroes can equip to attack Heroes or other minions.
    * Heroes represent the player; the player loses when the Hero's health reaches 0.
* **Set**: Cards are released in designated sets, which are usually based on a certain theme in the Warcraft universe.
    * The core game consists of two sets, Basic and Classic, which are always available in the game.
    * Expansion sets add newer cards to the game, and are released at regular intervals

### Approaching the Question

> Which are the most popular cards used in Ranked decks?

**Which are the most popular cards used in Ranked decks?**

We focus on the **Ranked** format where players get to decide which cards to include in their deck, therefore the cards' popularity are more accurately represented, and the gameplay is not subject to additional constraints that other game modes (like Tavern Brawls and Adventures) may impose.

**How do we determine popularity?**

We determine popularity by the number of decks that include *at least 1 copy* in the starting 30 cards (not generated by other effects). 

A deck can include **at most 2** copies of any card (1 for Legendary cards), thus a card's popularity is not heavily influenced by the number of copies players wish to use.

**Possible biases to consider**

* Decks and cards presented in this data may **not** necessarily be used by players in the game itself. However, without the means to track actual matches being played in the game itself, we instead use the list of decks submitted by players as an **approximation** of the decks and cards often used by the playerbase.

* Since **Neutral** cards can be used by multiple classes, they should be more popular than **Class** cards.

* For the Wild format, cards from the older sets may be more popular simply because they have been in the game longer.

* For the Standard format, cards from the **Basic** and **Classic** sets will be more popular because they do not rotate out of the format unlike expansion cards.

### Why address such a question?

If a certain card becomes too popular (i.e. the community thinks players must include it in their decks), it reduces the card variety in the metagame and makes gameplay frustrating for other players (amongst other consequences). In the long term, this may lead to player attrition and loss of potential revenue (when players purchase card packs or other cosmetics).

Historically, Blizzard has dealt with problematic cards in one of several ways:

* Introduce new cards that work well against them
* [Adjust the card's mechanics](https://hearthstone.gamepedia.com/Card_changes) to weaken their power
* Shift a card into the [Hall of Fame](https://hearthstone.gamepedia.com/Hall_of_Fame) which removes it from Standard format.

## Overview of Datasets used

There are three datasets we will use:

* `data.csv` contains a list of decks submitted by players to [HearthPwn](www.hearthpwn.com) from 2013 (pre-launch) to 2017,
* `refs.json` contains detailed information about all cards (both collectible and non-collectible) up to March 2017.
    * These files are taken from the following Kaggle webpage: [History of Hearthstone](https://www.kaggle.com/romainvincent/history-of-hearthstone/home).
* `cards_collectible.json` contains detailed information about all the cards that are collectible in the game.
    * This file can be obtained from [HearthstoneJSON](https://hearthstonejson.com/docs/cards.html).
    
### Loading the data

Assuming that the datasets are stored in a separate folder at the same subfolder level, the following code reads the datasets to be used:

```{r load data, message=FALSE}
data_path <- file.path("..", "data")

# data for decks
decks_raw <- read_csv(file.path(data_path, "data.csv"))
# data for collectible cards
cards_raw <- tbl_df(fromJSON(file.path(data_path, "cards_collectible.json"), flatten = TRUE))
# data for all cards
cards_all_raw <- fromJSON(file.path(data_path, "refs.json"), flatten = TRUE)
```

### Decks data

The first few rows and columns of the raw data `decks_raw` is shown below:

```{r echo=FALSE}
head(decks_raw[1:10], 6)
```

The `decks_raw` data has `r nrow(decks_raw)` rows and `r ncol(decks_raw)` columns. The first 11 columns `craft_cost` to `user` describe the deck's attributes (like date submitted, class, deck format) while the remaining 30 columns describe the cards each deck contains (based on the card's unique ID which can be referenced from the `cards_raw` data).

Detailed information on the variables can be found on the [Kaggle dataset](https://www.kaggle.com/romainvincent/history-of-hearthstone/home).

```{r}
names(decks_raw)
```

```{r}
glimpse(decks_raw) 
```

There are `r sum(!complete.cases(decks_raw))` rows that contain missing data. The missing values are in the 10th column, which contains the decks' `title` as submitted by the users, so they can be safely ignored.

```{r}
which(is.na(decks_raw), arr.ind = TRUE)
```

### Cards data

The first few rows and columns of the `cards_raw` data is shown below (some columns are shown truncated):

```{r}
col_trunc <- function(col, width, side = c("right", "left", "center"), 
                      ellipsis = "..."){
    if (class(col) == "character"){
        col <- str_trunc(col, width, side, ellipsis)
        return(col)
    } else {
        return(col)
    }
}

cards_raw[1:12] %>% 
    lapply(col_trunc, width = 10, side = "right") %>%
    as_tibble() %>% 
    head()
```

This dataset has `r nrow(cards_raw)` rows and `r ncol(cards_raw)` columns. The first 32 columns `artist` to `questReward` describe the characteristics of each card. Some of the characteristics, including `name`, Mana `cost`, `race`, `health` and `text`, can also been seen on the card itself.

Each card has two unique identifiers: a character/string `id` and an **integer `dbfId`**. The integer `dbfId` is used across both the deck and card datasets, which can be joined later. 

```{r}
names(cards_raw[1:32])
```

Some of the characteristics, including `name`, Mana `cost`, `race`, `health` and `text`, can also been seen on the card itself. Key `mechanics` are highlighted **bold** in the card text:

<img src = "../cardimg/825.png"  width="300">

The remaining columns are in fact nested under the `playRequirements` field, which specifies how certain cards can only be played in the game. These requirements are also implicitly/explicitly stated in the card text. (Column names are shown truncated)

```{r}
names(cards_raw)[33:65] %>% 
    str_trunc(25, "left")
```

Additional information on the variables can be found on [HearthstoneJSON](https://hearthstonejson.com/docs/cards.html).

The following card, based on its text, would require certain conditions to be played (a minion on the board, that has not taken any damage):

<img src = "../cardimg/180.png"  width="300">



The following computes the number of missing values in each field, with the exception of those that are present as lists or data frames (`mechanics`, `referencedTags`, `classes`, `entourage`). 

Many of the columns contain significant amounts of missing values. However, some key attributes that define the card such as `cardClass`, `collectible`, `dbfId`, `name`, `set`, and `type` do not contain any missing values.

```{r eval=FALSE}
# code is not run
sapply(cards_raw, function(x) sum(is.na(x)))
```


## Pre-processing

### Decks data

<!-- 
Filter for decks submitted after game launch
Split decks into attribute and composition(cards) data frames
Define factor columns
Add year and month variables for decks
Relabel deck format
-->

The raw dataset is split into two, `decks_attr` containing the deck attributes and `decks_comp` containing the deck composition (cards), joined by the `deck_id` identifier. We also exclude decks created before launch (there were many card changes in the alpha and beta stages, making card popularity very volatile).

The `decks_comp` data will be pivoted to long format later on, thus excluding fields that are not related to the cards will minimize the size of the dataset.

```{r include=FALSE, eval=FALSE}
# check for duplicate deck ids
sum(duplicated(decks_raw$deck_id))
```

```{r split deck data}
launch_date <- as.Date("2014-03-11")

decks_attr <- decks_raw %>% 
    filter(date >= launch_date) %>% 
    select(deck_id, craft_cost:deck_format, deck_set:user)

decks_comp <- decks_raw %>% 
    filter(date >= launch_date) %>% 
    select(deck_id, card_0:card_29)
```

Within `decks_attr`, the factor columns are identified and converted to the appropriate type.

```{r decks_attr factor}
fct_cols_attr <- c("deck_archetype", "deck_class", "deck_format", "deck_set", "deck_type")

decks_attr[fct_cols_attr] <- lapply(decks_attr[fct_cols_attr], factor)
rm(fct_cols_attr)
```

#### Years and Months

While each deck has a submission `date`, we may also be interested in grouping the decks by month (which corresponds to Ranked seasons) and by year (which is marked by expansion release dates instead of calendar dates).

Years in the game based on a time period that: 

* Starts with the release of the first [card set](https://hearthstone.gamepedia.com/Card_set) of each year, which usually falls around April.
* Ends with the release of the first card set the next year (non-inclusive). 

So based on the release dates (taken from the above link), the years would be:

* 2014-03-11 to 2015-04-01 (Live, Naxxramas, Goblin vs Gnomes)
* 2015-04-02 to 2016-04-25 (Blackrock, Grand Tournament, League of Explorers)
* 2016-04-26 to 2017-03-19 (Old Gods, Karazhan, Gadgetzan)

```{r decks_attr month and year}
# month (with year)
decks_attr$hsmonth <- as.yearmon(decks_attr$date)

# year (not by calendar)
decks_attr$hsyear <- case_when(
        decks_attr$date <= as.Date("2015-04-01") ~ "2014",
        decks_attr$date >= as.Date("2015-04-02") & 
            decks_attr$date <= as.Date("2016-04-25") ~ "2015",
        decks_attr$date >= as.Date("2016-04-26") ~ "2016"
    ) %>% 
        factor()
```

#### Deck Format

The Standard and Wild formats were formally introduced into the game on 2016-04-26 with the release of *Whispers of the Old Gods*; however the plot below shows that many decks from June 2014 to April 2016 were mistakenly labelled as Wild. 

```{r deck format plot, echo=FALSE}
decks_attr %>% 
    filter(deck_type == "Ranked Deck") %>% 
    ggplot(aes(x = as.yearmon(date), fill = deck_format)) + 
    geom_bar() +
    geom_vline(xintercept = as.yearmon("2016-04-26"), alpha = 0.4) +
    annotate("text", label = "Introduction of Standard/Wild formats", 
             x = as.yearmon("2016-04-26"), y = 16000,
             hjust = 1.01, angle = 0) +
    scale_x_yearmon() +
    scale_fill_discrete(labels = c("Standard", "Wild")) +
    labs(title = "Ranked Decks in Standard and Wild formats",
         x = "Month",
         y = "Number of Decks",
         fill = "Format") +
    theme(panel.grid.major.y = element_blank(), 
          plot.title = element_text(size = 12),
          axis.text.y = element_text(size = 9),
          panel.background = element_rect(fill = "#D7D7D7"))
```

We can simply relabel all decks created before 2016-04-26 as Standard since all cards before then are not separated by format:

```{r}
decks_attr$deck_format[decks_attr$date < as.Date("2016-04-26")] <- "S"
```

A summary of the processed data is shown below:

```{r, echo=FALSE}
glimpse(decks_attr)
```

```{r echo=FALSE}
summary(decks_attr)
```

### Cards data
<!--
Select core columns
Relabel card sets
Convert uppercase columns to titlecase
Define factor columns
Identify mislabelled cards; create list for re-labelling
-->

Many columns in the `cards_raw` data pertain to the card's stats, mechanics and play requirements, which is better explained by the text on the card image. So we choose to only include the columns that we consider are the core properties of the card:

* ID
* Name
* Class
* Rarity
* Type
* Card Set

```{r cards_simple}
cards_simple <- select(cards_raw,
                       dbfId,  name, cost,   cardClass, 
                       rarity, type,  set, collectible, id)
```

```{r, echo=FALSE}
glimpse(cards_simple)
```

The `set` column contains abbreviated names or nicknames and is not necessarily informative; we create a new column that uses the actual names of the card sets:

```{r}
unique(cards_simple$set)
```

```{r cards set}
# list for recoding card sets
cardset_lst <- list(
    "CORE" = "Basic",
    "EXPERT1" = "Classic",
    # 2014 sets
    "NAXX" = "Curse of Naxxramas",
    "GVG" = "Goblins vs Gnomes",
    # 2015 sets
    "BRM" = "Blackrock Mountain",
    "TGT" = "The Grand Tournament",
    "LOE" = "League of Explorers",
    # 2016 sets
    "OG" = "Whispers of the Old Gods",
    "KARA" = "One Night in Karazhan",
    "GANGS" = "Mean Streets of Gadgetzan",
    "HOF" = "Hall of Fame",
    # 2017 sets
    "UNGORO" = "Journey to Un'Goro",
    "ICECROWN" = "Knights of the Frozen Throne",
    "LOOTAPALOOZA" = "Kobolds & Catacombs",
    # 2018 sets
    "GILNEAS" = "The Witchwood",
    "BOOMSDAY" = "The Boomsday Project"
)

cards_simple$card_set <- recode(cards_simple$set, !!!cardset_lst,
                                .default = "Other")

# drop the set column to avoid confusion (it remains in the cards_raw dataset)
cards_simple$set <- NULL
# remove list (no longer needed)
rm(cardset_lst)
```

Some columns are entirely uppercase, which we convert to title case for readability:

```{r cards titlecase}
titlecase_cols <- c("cardClass", "rarity", "type")
cards_simple[titlecase_cols] <- lapply(cards_simple[titlecase_cols], 
                                       str_to_title)
rm(titlecase_cols)
```

The factor columns are then identified and converted to the appropriate type:

```{r cards factor}
fct_cols_cards <- c("cardClass", "rarity", "type", "card_set")

cards_simple[fct_cols_cards] <- lapply(cards_simple[fct_cols_cards], factor)
rm(fct_cols_cards)
```

A summary of the processed data is shown below:

```{r echo=FALSE}
summary(cards_simple)
```

### Mislabelled cards

Card IDs in the `decks_comp` data may be incorrect, due to the following reasons:

* Deck submissions are generated by user input.
* Users choose cards based on card names, not card IDs.
* However, there are cards that share the same name, but have different IDs.
    * Amongst each set of replicates, only one is collectible in-game.

Specifically, we are looking for the version of each card that is collectible (since all cards used in Ranked decks must be collectible). This step requires using the full card data `cards_all_raw` (which contains non-collectible cards).

The first step is to compile a list of unique card IDs found in `decks_comp`:

```{r}
# IDs of all cards used
cards_used  <- decks_comp %>%
        select(card_0:card_29) %>%
        unlist(use.names = FALSE) %>%  # flatten into a vector
        unique() %>%
        sort()
```

The following code filters for those IDs that are not found in our collectible cards data `cards_raw`:

```{r find missing ids}
# IDs of missing cards (using dbfId)
missing_cards <- cards_used[!cards_used %in% cards_raw$dbfId]
```

Using the above vector of IDs, we filter out the relevant non-collectible cards from the `cards_all_raw` dataset.

```{r locate missing cards}
# filter for those missing cards only
mssng_cards <- cards_all_raw %>% 
    select(dbfId, name, cardClass, type, collectible) %>% 
    filter(dbfId %in% missing_cards)
```

```{r echo=FALSE}
head(arrange(mssng_cards, name))
```


To find the correct IDs, we join the cards by name to the `cards_simple` dataset. The `dbfID.x` on the left would be replaced by the `dbfID.y` on the right:

```{r}
mislabelled <- mssng_cards %>% 
    select(dbfId, name) %>% 
    # also drops uncollectible mislabelled cards (which are not in simple_cards)
    inner_join(cards_simple, by = "name") %>% 
    arrange(name)
mislabelled
```

We create a named list that can be used within `recode()`:

```{r}
# list values are correct ids
mislab_recode <- as.list(mislabelled$dbfId.y)
# list names are mislabelled ids
names(mislab_recode) <- mislabelled$dbfId.x
```

The following step applies the `recode()` function to all the card columns `card_0` to `card_29` in our `decks_comp` data, which will relabel all the incorrect IDs with their respective IDs (`dbfID.x` with `dbfID.y` based on the table above).

```{r}
# relabel ids in data
decks_comp <- mutate_at(decks_comp, vars(starts_with("card")), 
                        recode, !!!mislab_recode)
```


```{r clean up workspace, message=FALSE}
# intermediate objects no longer needed
rm(cards_used, missing_cards, cards_all_raw, mssng_cards)
```

### Other objects
<!--
class colours
rarity colours
release dates of card sets
-->

The following objects are used for plotting and visualization:

```{r}
# class colours
class_colors <- c(
    "Druid" = "#FF7D0A",
    "Hunter" = "#569B56", #"#ABD473",
    "Mage" = "#40C7EB",
    "Paladin" = "#F58CBA",
    "Priest" = "#FFFFFF",
    "Rogue" = "#FFF569",
    "Shaman" = "#0070DE",
    "Warlock" = "#8787ED",
    "Warrior" = "#C79C6E",
    "Neutral" = "#777777"
)

# rarity colours
rarity_colors <- c(
    "Free" = NA,
    "Common" = "#000000",
    "Rare" = "#1E90FF",
    "Epic" = "#9932CC",
    "Legendary" = "#FFB90F"
)

# release dates of card sets
release_dates <- c(
    "Launch" = "2014-03-11", 
    "Naxxramas" = "2014-07-22", 
    "GvG" = "2014-12-08",
    "Blackrock" = "2015-04-02",
    "TGT" = "2015-08-24",
    "Explorers" = "2015-11-12",
    "Old Gods" = "2016-04-26",
    "Karazhan" = "2016-08-11",
    "Gadgetzan" = "2016-12-01"
)

# darker plot theme for this doc
theme_doc <- theme(panel.spacing = unit(1, "points"),
                   panel.grid.major.y = element_blank(), 
                   plot.title = element_text(size = 12),
                   axis.text.y = element_text(size = 9),
                   panel.background = element_rect(fill = "#C3C3C3"))
```

## Analysis of Popular Cards

The question we are trying to answer is

> Which are the most popular cards used in Ranked decks (Standard format)?

We measure card popularity by counting the number of decks that include at least one copy of that card.

### Preparation and Validation

To begin, we filter only for *Ranked Decks in Standard format*, pick out the variables of interest, join the decks to their compositions, and pivot the card columns.

```{r rs_decks_all }
# create set of ranked standard decks for pct calculation
rs_decks_all <- decks_attr %>% 
    filter(deck_type == "Ranked Deck",
           deck_format == "S")
```

```{r attr-comp join and pivot}
ranked_deck_cards <- rs_decks_all %>%
    select(deck_id, deck_class, 
           date, hsmonth, hsyear, 
           deck_format, deck_set, craft_cost) %>% 
    
    inner_join(decks_comp, by = "deck_id") %>% 
    gather(card_no, card_id, card_0:card_29) %>% 
    select(-card_no) %>% # drop the key column which is unnecessary
    
    # select unique combinations of deck and card (ignore duplicates)
    distinct(deck_id, card_id, .keep_all = TRUE)
```

```{r}
glimpse(ranked_deck_cards)
```



### Popular Cards of All Time

To derive overall card popularity, we count the number of times each `card_id` appears, then join it to the `cards_simple` data to retrieve the card details. After that, we sort by number of decks. In addition, we also calculate the percentage of all Ranked Standard decks that feature each card.

```{r pop cards all}
pop_cards_all <- ranked_deck_cards %>% 
    group_by(card_id) %>% 
    summarise(n_decks = n(),
              pct_all = n_decks * 100 / nrow(rs_decks_all)) %>% 
    inner_join(cards_simple, by = c("card_id" = "dbfId"))
```

As the table below shows, the most popular cards are expected to be **Neutral** class since they can be used by all nine classes in the game, as opposed to Class cards:

```{r echo=FALSE}
pop_cards_all %>%
    select(name, cardClass, rarity, card_set, n_decks, pct_all) %>% 
    arrange(desc(n_decks)) %>% 
    slice(1:10) %>% 
    kable(digits = 1, caption = "Most Popular Cards (by number of decks)",
          col.names = c("Name", "Class", "Rarity", "Card Set", 
                        "No. of Decks", "% of All Decks"))
```

```{r include=FALSE, eval=FALSE}
# retrieve dbfids for images
pop_cards_all %>% 
    select(name, card_id, n_decks) %>% 
    arrange(desc(n_decks)) %>% 
    slice(1:10)
```


<img src = "../cardimg/825.png"  width="150">
<img src = "../cardimg/1721.png" width="150">
<img src = "../cardimg/2262.png" width="150">
<img src = "../cardimg/2078.png" width="150">
<img src = "../cardimg/1793.png" width="150">
<img src = "../cardimg/763.png" width="150">
<img src = "../cardimg/374.png" width="150">
<img src = "../cardimg/2064.png" width="150">
<img src = "../cardimg/1659.png" width="150">
<img src = "../cardimg/749.png" width="150">

**Which abilities/properties makes these cards popular?**

* *Card draw* - Having more cards in hand gives the player more options to try and get an advantage over the opponent.
* *Deathrattle* - Triggers additional effects when the minion is removed from the board.
* *Enhancement* - Improving the player's other cards on the board/in hand.
* *Taunt* - Protects the player's hero and minions from attack damage.

As a side note, the Hall of Fame was only created in April 2017; the popularity of those three cards (Azure Drake, Sylvanas Windrunner, Ragnaros the Firelord) may have been one of the factors behind Blizzard's decision to move them out of Standard format and into the Hall of Fame.

The following plot shows the relative popularity of all cards in the dataset:

```{r pop cards all plot, echo = FALSE, fig.height=4}
pop_cards_all %>% 
    top_n(10, n_decks) %>% 
    mutate(name = reorder(name, n_decks, median)) %>% 
    
    ggplot(aes(x = name, y = n_decks)) +
    geom_bar(stat = "identity") +
    geom_point(aes(color = rarity, y = 1000), shape = 18, size = 3) +
    scale_colour_manual(values = rarity_colors) +
    coord_flip(expand = FALSE, ylim = c(0, 63000), xlim = c(0.4, 10.6)) +
    labs(title = "Most Popular Cards", 
         x = "", y = "No. of Decks", color = "Card Rarity") +
    theme_doc +
    theme(legend.position = "bottom")
```

### Popular Class Cards

Since Neutral cards are expected to be more popular in general, we follow up by asking:

> Amongst Class cards, which is the most popular card for each class?

By tallying the total number of decks by class, we can evaluate for each class the percentage of decks which use the most popular class cards.

```{r decks class}
rs_decks_class <- rs_decks_all %>% 
    group_by(deck_class) %>% 
    summarise(tot_decks_class = n())
```

```{r pop cards class, warning=FALSE}
pop_cards_class <- pop_cards_all %>% 
    filter(!is.na(cardClass), cardClass != "Neutral") %>% 
    select(cardClass, name, rarity, card_set, n_decks) %>% 
    inner_join(rs_decks_class, by = c("cardClass" = "deck_class")) %>% 
    mutate(pct_class = n_decks * 100 / tot_decks_class)
```

The most popular class cards are as follows:

```{r echo=FALSE, warning=FALSE}
pop_cards_class %>% 
    group_by(cardClass) %>%
    top_n(1, n_decks) %>% 
    ungroup() %>% 
    arrange(cardClass, desc(n_decks)) %>% 
    kable(digits = 1, caption = "Most Popular Class-Specific Cards",
          col.names = c("Class", "Name", "Rarity", "Card Set", 
                        "No. of Decks", "Total No. of Class Decks", "% of Class Decks"))
```

```{r include=FALSE, eval=FALSE}
# retrieve dbfids for images
pop_cards_all %>% 
    filter(!is.na(cardClass), cardClass != "Neutral") %>% 
    group_by(cardClass) %>% 
    top_n(1, n_decks) %>% 
    ungroup() %>% 
    arrange(cardClass, desc(n_decks)) %>% 
    select(name, card_id)
```


<img src = "../cardimg/64.png" width="150">
<img src = "../cardimg/437.png" width="150">
<img src = "../cardimg/662.png" width="150">
<img src = "../cardimg/847.png" width="150">
<img src = "../cardimg/1363.png" width="150">
<img src = "../cardimg/904.png" width="150">
<img src = "../cardimg/766.png" width="150">
<img src = "../cardimg/846.png" width="150">
<img src = "../cardimg/401.png" width="150">

**Which abilities/properties makes the above cards popular?**

* *Targeted damage* - Being able to remove problematic minions that are otherwise protected by Taunt, or when your own minions do not have the damage to kill it.
* *Transformation/removal* - Alternatively, simply removing such minions from the board might be more feasible.
* *Weapons* - Most minions cannot attack the turn they are played, but weapons enable the player to attack immediately.

Some of the other popular cards (in the following plot) also have effects like *card draw* or *area of effect*.

While Frostbolt appeared in the largest number of decks, many of the most popular cards were included in more than 90% of decks from their class. The only exception would be Power Overwhelimg (Warlock) which only appeared in 63% of Warlock decks.

As the table shows, cards from the **Basic** and **Classic** sets tend to be more popular because they do not rotate out of Standard format, and they have been around since the launch of Hearthstone.

The following plot shows the relative popularity of the top 3 cards for each class. Dashed lines indicate the total number of decks representing each class in the dataset:

```{r pop cards class plot, echo=FALSE, fig.height=7, warning=FALSE}
pop_cards_class %>% 
    group_by(cardClass) %>% 
    top_n(3, n_decks) %>% 
    ungroup() %>%  # ungroup is required before reordering chars/factors
    mutate(name = reorder(name, n_decks, median)) %>% 
    
    ggplot(aes(x = name, y = n_decks, fill = cardClass)) +
    geom_bar(stat = "identity", alpha = 1) +
    geom_hline(aes(yintercept = tot_decks_class), linetype = 2, size = 0.7) +
    geom_point(aes(color = rarity, y = n_decks + 500), shape = 18, size = 3) +
    scale_colour_manual(values = rarity_colors) +
    scale_fill_manual(values = class_colors, guide = FALSE) +
    coord_flip(expand = FALSE, xlim = c(0.4, 3.6),
               ylim = c(0, max(pop_cards_class$tot_decks_class) + 1000)) +
    facet_wrap(vars(cardClass), ncol = 1, scales = "free_y", strip.position = "right") +
    labs(title = "Most Popular Class-Specific Cards in Ranked Standard Decks", 
         x = "", y = "No. of Decks", color = "Card Rarity") +
    theme_doc +
    theme(legend.position = "bottom")
```

### Popular Cards by Year

Our `decks_attr` data spans 3 years from March 2014 to March 2017. Recall that years are defined by release dates of card sets instead of calendar months:

* 2014-03-11 to 2015-04-01 (Live, Naxxramas, Goblin vs Gnomes)
* 2015-04-02 to 2016-04-25 (Blackrock, Grand Tournament, League of Explorers)
* 2016-04-26 to 2017-03-19 (Old Gods, Karazhan, Gadgetzan)

To derive card popularity, we count the number of times each `card_id` appears in each `hsyear`:

```{r decks year}
rs_decks_year <- rs_decks_all %>% 
    group_by(hsyear) %>% 
    summarise(tot_decks_year = n())
```

```{r pop cards year}
pop_cards_year <- ranked_deck_cards %>% 
    group_by(hsyear, card_id) %>% 
    summarise(n_decks = n()) %>% 
    inner_join(cards_simple, by = c("card_id" = "dbfId"))
```

The following table shows the top 3 cards for each year. As expected, Neutral cards are used more often than class-specific cards. At the same time, the most popular card in 2015 and 2016 (Dr. Boom and Azure Drake) were considered even more essential than the others, considering that they were included in more than 40% of decks submitted:

```{r, echo=FALSE, results='asis'}
pop_cards_year %>% 
    select(hsyear, name, rarity, card_set, n_decks) %>% 
    group_by(hsyear) %>% 
    top_n(3, n_decks) %>% 
    ungroup() %>% 
    inner_join(rs_decks_year, by = "hsyear") %>% 
    mutate(pct_year = n_decks * 100 / tot_decks_year) %>% 
    arrange(hsyear, desc(n_decks)) %>% 
    kable(digits = 1, caption = "Top 3 Cards in the Years 2014, 2015, 2016",
          col.names = c("Year", "Name", "Rarity", "Card Set",
                        "No. of Decks", "Total No. of Decks", "% of Decks"))
```

```{r include=FALSE, eval=FALSE}
# retrieve dbfids for images
pop_cards_year %>% 
    group_by(hsyear) %>% 
    top_n(3, n_decks) %>% 
    ungroup() %>% 
    arrange(hsyear, desc(n_decks)) %>% 
    select(hsyear, name, card_id)
```

<img src = "../cardimg/1793.png" width="150">
<img src = "../cardimg/763.png" width="150">
<img src = "../cardimg/778.png" width="150">
<img src = "../cardimg/2078.png" width="150">
<img src = "../cardimg/2064.png" width="150">
<img src = "../cardimg/825.png" width="150">
<img src = "../cardimg/1721.png" width="150">
<img src = "../cardimg/2949.png" width="150">

**Which abilities/properties makes the above cards popular?**

* *Battlecry* - Triggers an effect when the card is played. **Brann Bronzebeard** complements such cards very well.
* *Deathrattle* - Triggers an effect when the minion is killed.
* *Taunt* - Protects the hero and minions from attack damage.

The following plot illustrates the relative popularity of cards in each year, and clearly shows the popularity of 2015 Dr. Boom and 2016 Azure Drake over the others.

```{r pop cards year plot, echo=FALSE}
pop_cards_year %>% 
    group_by(hsyear) %>% 
    top_n(5, n_decks) %>% 
    ungroup() %>% 
    # rearrange cards in alphabetical order from top to bottom
    mutate(name = factor(name)) %>% 
    mutate(name = factor(name, levels = rev(levels(name)))) %>% 
    
    ggplot(aes(x = name, y = n_decks)) +
    geom_bar(stat = "identity") +
    geom_point(aes(color = rarity, y = 1000), shape = 18, size = 3) +
    scale_colour_manual(values = rarity_colors) +
    facet_wrap(vars(hsyear), ncol = 1, scales = "free_y", strip.position = "right") +
    coord_flip(expand = FALSE, ylim = c(0, max(pop_cards_year$n_decks) + 1000),
               xlim = c(0.4, 5.6)) +
    labs(title = "Most Popular Cards in Ranked Standard Decks every year (Apr-Mar)",
         x = "", y = "No. of Decks") +
    theme_doc
```


### Popular Cards by Month

On the other hand, the Ranked ladder resets to mark a new Season at the start of every calendar month. Players take this opportunity to improve upon their previous ranking, and they may try to include new cards to overcome the popular decks that they tend to encounter.

```{r echo=FALSE}
decks_attr %>% 
    filter(deck_type == "Ranked Deck", 
           deck_format == "S") %>% 
    ggplot(aes(x = hsmonth)) +
    geom_bar(stat = "count") +
    scale_x_yearmon() +
    geom_vline(xintercept = as.yearmon(release_dates), alpha = 0.2) +
    annotate("text", label = names(release_dates), 
             x = as.yearmon(release_dates), y = 16000,
             hjust = 1, angle = 60) +
    labs(title = "Standard Ranked Decks Submitted Every Month",
         x = "Month", y = "")
```

From the graph above, there is usually a spike in the number of decks submitted when a new card set is released (marked by the vertical lines). The number of submissions gradually tapers off until the next card set is released.

A possible question to ask may be:

> Are there any shifts in card popularity in the three months following the release of the Old Gods card set (2016-04-26)?

The plot below charts the number of decks submitted every day for the first three months following 26th April. The number of decks submitted (and the number of cards included) sharply tapers off after the first week. 

```{r month plot, echo=FALSE}
decks_attr %>% 
    filter(date >= as.Date("2016-04-26"),
           date <= as.Date("2016-07-25"),
           deck_type == "Ranked Deck",
           deck_format == "S") %>% 
    ggplot(aes(x = date)) +
    # geom_point(stat = "count", size = 1) +
    geom_line(stat = "count") +
    scale_x_date(date_breaks = "2 weeks") +
    labs(title = "Number of Decks Submitted (Apr 26 to Jul 25)", 
         x = "Date", y = "")
```

The following code evaluates the card popularity for each month from April 2016 to July 2016, as well as the percentage of decks that included these cards in each month:

```{r pop cards month}
pop_cards_2016_aprjul <- ranked_deck_cards %>%
    filter(date >= as.Date("2016-04-26"),
           date <= as.Date("2016-07-25")) %>% 
    group_by(hsmonth, card_id) %>% 
    summarise(n_decks = n()) %>% 
    inner_join(cards_simple, by = c("card_id" = "dbfId"))
```

```{r decks month}
rs_decks_2016_aprjul <- rs_decks_all %>% 
    filter(date >= as.Date("2016-04-26"),
           date <= as.Date("2016-07-25")) %>% 
    group_by(hsmonth) %>% 
    summarise(tot_decks_month = n())
```


While cards from the latest card set were popular in the first week (still in April), cards from the older card sets (such as Classic) returned to the top spots in subsequent months. However, since the data concerns decks submitted (not necessarily played in games) we cannot infer that the new cards fell in popularity, only that decks submitted in the later weeks favoured including older cards over the new ones.

```{r echo=FALSE, results='asis'}
pop_cards_2016_aprjul %>% 
    select(hsmonth, name, card_set, n_decks) %>% 
    group_by(hsmonth) %>% 
    top_n(5, n_decks) %>% 
    inner_join(rs_decks_2016_aprjul, by = "hsmonth") %>% 
    mutate(pct_month = n_decks * 100 / tot_decks_month) %>% 
    arrange(hsmonth, desc(n_decks)) %>% 
    kable(digits = 1, caption = "Popular Cards in Apr-Jul 2016 following Old Gods release",
          col.names = c("Month", "Name", "Card Set", "No. of Decks", 
                        "Total No. of Decks", "% of Decks"))
```



### Popular Cards by Card Set

> Which were the most popular cards (Neutral and Class-specific) for each card set?

The first table shows the most popular Neutral card from each card set, with many of them being Legendary:

```{r pop cards set, echo=FALSE, results='asis'}
pop_cards_all %>% 
    filter(cardClass == "Neutral") %>% 
    group_by(card_set) %>% 
    top_n(1, n_decks) %>% 
    arrange(card_set, desc(n_decks)) %>% 
    select(card_set, name, cardClass, rarity, n_decks) %>% 
    kable(caption = "Most Popular Neutral Cards by Card Set",
          col.names = c("Card Set", "Name", "Class", "Rarity", "No. of Decks"))
```

```{r include=FALSE, eval=FALSE}
# retrieve dbfIds for images
pop_cards_all %>% 
    filter(cardClass == "Neutral") %>% 
    group_by(card_set) %>% 
    top_n(1, n_decks) %>% 
    arrange(card_set) %>% 
    `$`(card_id)
```

<img src = "../cardimg/906.png" width="150">
<img src = "../cardimg/2262.png" width="150">
<img src = "../cardimg/763.png" width="150">
<img src = "../cardimg/1793.png" width="150">
<img src = "../cardimg/2078.png" width="150">
<img src = "../cardimg/825.png" width="150">
<img src = "../cardimg/2949.png" width="150">
<img src = "../cardimg/40408.png" width="150">
<img src = "../cardimg/40596.png" width="150">
<img src = "../cardimg/39941.png" width="150">
<img src = "../cardimg/2736.png" width="150">
<img src = "../cardimg/38496.png" width="150">

The following plot shows the relative popularity of Neutral cards in each card set. Notably, the Basic, Blackrock Mountain and Karazhan sets have an overwhelming favourite (with a large difference over the 2nd most popular card):

```{r pop cards set plot, echo=FALSE, warning=FALSE, fig.height=6, fig.width=9}
pop_cards_all %>% 
    filter(cardClass == "Neutral") %>% 
    group_by(card_set) %>% 
    top_n(4, n_decks) %>% 
    ungroup() %>% 
    mutate(name = reorder(name, n_decks, median)) %>% 
    
    ggplot(aes(x = name, y = n_decks)) +
    geom_bar(stat = "identity") +
    geom_point(aes(color = rarity, y = 0), shape = 18, size = 3) +
    scale_colour_manual(values = rarity_colors) +
    scale_y_continuous(minor_breaks = NULL) +
    facet_wrap(vars(card_set), ncol = 3, scales = "free") +
    coord_flip() +
    labs(title = "Relative Popularity of Neutral Cards in each Card Set",
         x = "", y = "No. of Decks", color = "Card Rarity") +
    theme_doc +
    theme(strip.background = element_rect(fill = NA),
          axis.text.x = element_text(size = 8),
          axis.text.y = element_text(size = 7),
          legend.position = "bottom")
```

The next table shows the most popular Class-specific cards from each card set. 

```{r pop cards set class, echo=FALSE, results='asis'}
pop_cards_all %>% 
    filter(cardClass != "Neutral") %>% 
    group_by(card_set) %>% 
    top_n(1, n_decks) %>% 
    arrange(card_set, desc(n_decks)) %>% 
    select(card_set, name, cardClass, type, rarity, n_decks) %>% 
    kable(caption = "Most Popular Class Cards by Card Set",
          col.names = c("Card Set", "Name", "Class", "Type", "Rarity", "No. of Decks"))
```

```{r include=FALSE, eval=FALSE}
# retrieve dbfIds for images
pop_cards_all %>% 
    filter(cardClass != "Neutral") %>% 
    group_by(card_set) %>% 
    top_n(1, n_decks) %>% 
    arrange(card_set) %>% 
    `$`(card_id)
```

<img src = "../cardimg/662.png" width="150">
<img src = "../cardimg/2288.png" width="150">
<img src = "../cardimg/904.png" width="150">
<img src = "../cardimg/1805.png" width="150">
<img src = "../cardimg/2026.png" width="150">
<img src = "../cardimg/846.png" width="150">
<img src = "../cardimg/3015.png" width="150">
<img src = "../cardimg/40938.png" width="150">
<img src = "../cardimg/39712.png" width="150">
<img src = "../cardimg/2610.png" width="150">
<img src = "../cardimg/38267.png" width="150">

The following plot shows the relative popularity of Class cards in each card set. Besides the Hall of Fame (which usually consists of Classic cards), the Naxxramas set has an overwhelming favourite in Death's Bite (Warrior), which was highly valued both for its strong damage and Deathrattle effect. 

```{r pop cards set class plot, echo=FALSE, fig.height=6.5, fig.width=9}
pop_cards_all %>% 
    filter(cardClass != "Neutral") %>% 
    group_by(card_set) %>% 
    top_n(4, n_decks) %>% 
    ungroup() %>% 
    mutate(name = reorder(name, n_decks, median)) %>% 
    
    ggplot(aes(x = name, y = n_decks, fill = cardClass)) +
    geom_bar(stat = "identity") +
    # geom_point(aes(color = rarity, y = 0), shape = 18, size = 3) +
    # scale_colour_manual(values = rarity_colors) +
    scale_fill_manual(values = class_colors) +
    scale_y_continuous(minor_breaks = NULL) +
    facet_wrap(vars(card_set), ncol = 3, scales = "free") +
    coord_flip() +
    labs(title = "Relative Popularity of Class Cards in each Card Set",
         x = "", y = "No. of Decks", fill = "Class") +
    theme_doc +
    theme(strip.background = element_rect(fill = NA),
          axis.text.x = element_text(size = 8),
          axis.text.y = element_text(size = 7),
          legend.position = "bottom")
```



### Popular Cards by Mana Cost

> Between 1 and 10-cost mana cards, which cards were the most popular amongst cards of the same cost?

* Given that players can only have a maximum of 10 mana per turn, we only consider cards with a default mana cost of **0 to 10** (cards with mana costs higher than 10 require additional conditions to reduce their cost).

```{r echo=FALSE, results="asis"}
pop_cards_all %>%
    filter(cost <= 10) %>%
    group_by(cost) %>%
    top_n(1, n_decks) %>%
    select(cost, name, cardClass, type, card_set, n_decks) %>%
    arrange(cost) %>%
    kable(caption = "Most Popular Class Cards by Mana Cost",
          col.names = c("Mana Cost", "Name", "Class", "Type", "Card Set", "No. of Decks"))
```


```{r include=FALSE}
pop_cards_all %>%
    filter(cost <= 10) %>%
    group_by(cost) %>%
    top_n(1, n_decks) %>%
    ungroup() %>% 
    arrange(cost) %>%
    select(cost, name, card_id)
```

<img src = "../cardimg/254.png" width="150">
<img src = "../cardimg/613.png" width="150">
<img src = "../cardimg/749.png" width="150">
<img src = "../cardimg/1659.png" width="150">
<img src = "../cardimg/763.png" width="150">
<img src = "../cardimg/825.png" width="150">
<img src = "../cardimg/1721.png" width="150">
<img src = "../cardimg/2078.png" width="150">
<img src = "../cardimg/374.png" width="150">
<img src = "../cardimg/1186.png" width="150">
<img src = "../cardimg/38496.png" width="150">

The following plot shows the relative popularity of cards with the same mana cost. Notably, the groups of 5, 7, 8 and 9-cost cards have the most popular cards leading with a significant margin. There are also some Class cards that fare well against cards with the same mana cost, despite our expectation that Neutral cards should be favoured over Class cards.

```{r echo=FALSE, fig.height=7, fig.width=9}
pop_cards_all %>%
    filter(cost <= 10) %>%
    group_by(cost) %>%
    top_n(5, n_decks) %>%
    ungroup() %>% 
    mutate(name = reorder(name, n_decks, median)) %>% 
    
    ggplot(aes(x = name, y = n_decks, fill = cardClass)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = class_colors) +
    facet_wrap(vars(cost), scales = "free", ncol = 3) +
    coord_flip() +
    labs(title = "Relative Popularity of Cards grouped by Mana Cost",
         x = "", y = "No. of Decks", fill = "Class") +
    theme_doc +
    theme(strip.background = element_rect(fill = NA),
          axis.text.x = element_text(size = 8),
          axis.text.y = element_text(size = 7),
          legend.position = "bottom")
```


## Summary and Reflection

So far, we have looked at cards that users tend to include in decks in Standard format for Ranked play, which is also used for official Hearthstone tournaments - making these popular cards highly visible to a wide audience. We have also looked at card popularity when broken down by various categories, such as class, time period and card set.

**Are there any limitations to the data that may have affected our analysis?**

The major limitation of this data is that it only looks at decks submitted to a third-party website, which brings up the following issues:

* Decks submitted may not necessarily be played in the game itself, either because other players think it is too weak, or because the user submits a joke deck that contains absurd combinations of cards and is not meant to be taken seriously.
* There is no data on how often decks and cards are **actually played** in the Ranked format.
* Likewise, there is no information on how effective the decks are at **winning games**. While the `rating` attribute may reflect how strong other players consider a deck, it is also biased towards the popularity of the user as well as the date of submission:
    * A deck may be initially strong and highly rated, but as new cards are introduced and old cards are removed from Standard format, the deck may wane in strength, but users are unlikely to retract their votes by this point in time.
    
**How can we expand on this analysis?**

* Examine popular combination of cards that complement each other well.
* Examine whether the crafting cost (in dust) of decks has any relation to its popularity (rating).